Multisequence algorithm for coarse-grained biomolecular simulations: application to protein folding switching

### 2017-05-23 Manuscript.tex
changes220517

\begin{abstract}
We evaluate a generalized-ensemble algorithm for coarse-grained simulations of biomolecules that allows the thermodynamics of multiple sequences to be calculated in a single ``multisequence" run. By carrying out a random walk in sequence space, conformational sampling is also enhanced. Escape from local energy minima is achieved by visiting sequences for which the minima is shallower or absent. We apply the algorithm to a coarse-grained model for protein folding with 7 atoms per amino acid and apply it to sets of sequences with up to 35 amino acids spanning sequence space  With the multisequence algorithm, we show that the folding thermodynamics of $>$1,000 sequences at relatively low temperatures can be readily calculated. We use the resulting data to explore the statistics of pathways in sequence space between the two pairs of folds.

@@

\section{Introduction}
\noindent
Modern investigations into biophysical properties of biomolecular systems are routinely carried out not only on a single species but on multiple molecular variants.~\cite{Vukmirovic2000,Nickson2010} For example, protein engineering methods are used to provide mechanistic insight into the specificity of protein-protein and protein-DNA interactions, to compare the  conformational propensities of disease-related mutations of aggregation-prone proteins, and optimize biomolecular properties through directed evolution. Moreover, large-scale screening techniques such as phage display and microarray chips are used to provide biophysical properties across entire genomes. In short, through the study of a biomolecular system and related variants molecular variants are studied in order to gain a more complete picture of the biophysical and biochemical properties of the system and provide evolutionary context.

@@

We first apply the MS algorithm to a set of 144 model sequences with 16 amino acids and compare with the simulated tempering (ST) algorithm. We find that the computational efficiency of the MS algorithm is never less efficient than ST even at the lowest studied temperature. The model folds spontaneously designed sequences into stable native states with diverse protein-like sequences, and captures features such as mutational robustness. It therefore provide a biophysical basis for the sequence-structure and sequence-stability relationships of proteins at a conceptual level. This model has been used, e.g., to study the sequence-structure relationship of proteins in the "marshlands" between protein folds. Experimental studies have indicated that single point mutations are capable to take change one protein fold into another. The multisequence algorithm allow us to characterize the stability of and structure of all 1,024 sequence in the binary space between the two folds. The implication of these results for the fold switching and evolution is discussed. \\

@@

While the 16-amino acid A1 and N1 sequences fold into $\alpha$-helical and $\beta$-hairpin folds, respectively, the two 35-amino acid sequences A2 and TA adopt the folds denoted IIA and IIB, as shown in Fig.~X.

The large number of sequences considered allows us to carry out a much more systematic analysis than previously~\cite{Holzgrafe2014,Holzgrafe2015} of the sequence-structure-stability relationship in the sequence region that connect these two folds pairs. We find that...

@@

\subsection{Generalized-ensemble algorithms and simulated tempering}
\noindent
Conventional Monte Carlo simulations of the canonical distribution is problematic at low temperatures $T$ for many physical systems because simulations tend to become trapped in local energy energy minima and hamper representative sampling of configurational space. The basic idea of generalized-ensemble algorithms~\cite{Mitsutake2001} is to alleviate this trapping problem by sampling states using a non-Boltzmann weight factor and/or expand the state space with additional dynamical parameters, such that a more efficient random walk in potential energy can be achieved.

@@

In analogy with ST, the MS algorithm simulates the joint probability distribution
\begin{equation}
P(\sbar,\rbar) =\dfrac{1}{\tilde{Z}}e^{-\beta E(\sbar,\rbar) + h(\sbar)}\,, 
\label{MS}
\end{equation}
where  
\begin{equation}
\tilde{Z} = \sum_{\sbar}\sum_{\rbar} e^{-\beta E(\sbar,\rbar)+ h(\sbar)}\,
\end{equation}
and the first sum goes over a set of allowed sequences $\sbar$. The simulation parameters $h(\sbar)$, similarly to the parameters $g_m$ in ST, control the marginal distribution $P(\sbar)=\tilde{Z}^{-1}\sum_{\rbar} e^{-\beta E(\sbar,\rbar)+ h(\sbar)} = \tilde{Z}^{-1}e^{-\beta F(\sbar)+ h(\sbar)}$ and a roughly flat $P(\sbar)$ can be achieved by choosing $h(\sbar) \approx \beta F(\sbar)$, where $F(\sbar)$ is the free energy of sequence $\sbar$ at temperature $T$.

@@

\section{Model and Methods}
\subsection{Coarse-grained 3-letter model for protein folding}
\noindent
All calculations were carried out using the coarse-grained model for protein folding developed in Ref.~\citenum{Bhattacherjee2012}. In this model, there are 3 different amino acid types: hydrophobic (h), polar (p) and turn-type (t). The backbone chain is represented atomistically by the N, H, $\mathrm{C}_\alpha$, $\mathrm{H}_{\alpha 1}$, C$'$ and O atoms. By contrast, the sidechain represention is simplified to a single enlarged $\mathrm{C}_\beta$ atom, which is geometrically identical for h and p types. The sidechain is absent for the t type which instead has an $\mathrm{H}_{\alpha 2}$ atom. The t type is therefore closely related to glycine. All bond lengths, bond angles, and peptide plane angles (180$^\circ$) are held fixed. Hence, an $N$-amino acid chain conformation $\rbar$ can, for any sequence $\sbar$, is therefore described by the set of 2$N$ backbone torsional angles $\{\phi_i$, $\psi_i\}_{i=1}^{N}$.

This geometrical description is paired with a simplified but finely tuned energy function with 4 terms: $E= E_\mathrm{ev} + E_\mathrm{loc} + E_\mathrm{hb} + E_\mathrm{hp}$. The first two, $E_\mathrm{ev}$ and $E_\mathrm{loc}$, represent excluded-volume effects of all atoms and local electrostatic effects, respectively. The hydrogen-bond energy, $E_\mathrm{hb}$, represents directionally dependent interactions between NH and CO groups which are necessary for secondary structure formation. Finally, the ``hydrophobicity" term, $E_\mathrm{hp}$, implements pairwise Lennard-Jones-like interactions between the $C_\beta$ atoms of h amino acids which are necessary for driving chain collapse during folding. Various model parameters, e.g., the strengths of hydrophobic attractions and hydrogen bonding, were determined based on the ability of the model to spontaneously fold a set of model sequences with 18-54 amino acids into  structurally diverse and thermodynamically stable native states with both $\beta$ and $\alpha$-structure. As it turned out, this strategy made the model robust enough to fold sequences designed to have mised $\alpha$ and $\beta$ structure.

@@

\begin{table}
\caption{\label{tab1} List of model sequences of different lengths $N$ studied in this work.}
\begin{ruledtabular}
\begin{tabular}{lcr}
Name & $N$ & Sequence\footnote{In addition, we study two sets of 1,024 sequences derived from the A1-N1 and A2-TN pairs (see text) denoted $\mathrm{S16}_{1024}$  and $\mathrm{S35}_{1024}$, respectively, and a restricted set $\mathrm{S16}_{144}$ with 144 sequences taken from S16 ($\mathrm{S16}_{144}$ was also studied in Ref.~\protect\citenum{Holzgrafe2014}). } \\
\hline
A1 & 16 & pphpphhpphpphhpp \\
N1 & 16 & phphphpttphphphp \\
R1 & 16 & \textbf{(add)}\\
R2 & 16 &  \textbf{(add)}\\
A2 & 35 & (A1)ttt(A1)\\
TN & 35 & (A1)ttt(N1)\\
\end{tabular}
\end{ruledtabular}
\end{table}

\subsection{Monte Carlo simulation parameters}
\noindent
Both ST and MS simulations are carried out with two types of conformational updates $\rbar\rightarrow\rbar'$: (1) a global pivot move (20\%) which randomly picks a $\phi_\mathrm{i}$ angle or $\psi_\mathrm{i}$ angle and assigned a new value between $-\pi$ and $\pi$; and (2) a semi-local move (80\%) which turns the $\phi_\mathrm{i}$ and $\psi_\mathrm{i}$-angles of 4 consecutive amino acids in a coordinated manner.~\cite{Favrin2001} In MS simulations, sequence updates $\sbar\rightarrow\sbar'$ are carried out by randomly picking a new sequence $\sbar'\ne\sbar$ and applying the accept-reject criterion in Eq.~\ref{accrej}. A sequence move is attempted every $1,000$ MC step while temperature updates $m\rightarrow m'$ are attempted every 100 steps. The simulations carried out in this work are summarized in Tables~II.

\begin{table}
\caption{\label{tab2} List of simulations carried out in this work. }
\begin{ruledtabular}
\begin{tabular}{lccccr}
Runs & Algorithm & $\kb T$  & MC steps &  Sequences\\
\hline
32 & ST & 0.43--0.65 & $8\times 10^6$ &A1\\ 
32 & ST & 0.43--0.65 & $8\times 10^6$ &N1\\ 
32 & ST & 0.43--0.65 & $8\times 10^6$ &R1\\ 
32 & ST & 0.43--0.65 & $8\times 10^6$ &R2\\ 
32$\times$8\footnote{32  runs per temperature for 8 temperatures in the range given.} & MS &0.43--0.65& $144\times 10^6$ & $\mathrm{S16}_{144}$\\
16 & MS & 0.43  & $5\times 10^9$ &  $\mathrm{S16}_{1024}$ \\
16 & MS & 0.46 & $4\times 10^9$ &  $\mathrm{S35}_{1024}$ \\
\end{tabular}
\end{ruledtabular}
\end{table}

@@

For folds A and B, $Q$ is determined by defining a set of backbone-backbone hydrogen bonds that are present in those folds (ignoring N and C terminal amino acids). For fold A, the fold-specific set of contacts (i,j) between the CO group of amino acid i and the NH group of amino acid j is $\{$(2,6), (3,7), (4,8), (5,9), (6,10), (7,11), (8,12), (9,13), (10,14), (11,15)$\}$, and for fold B the contact set is $\{$(3,14), (5,12), (7,10), (10,7), (12,5), (14,3)$\}$. A hydrogen bond is considered formed if (i) the NHO and COH angles both are $>90^\circ$ and (ii) the HO distance is $<2.7$~{\AA}.

@@

\begin{figure}
\includegraphics[width=6.0cm]{MCtraj}
\caption{(A) Example simulation trajectory of the MS algorithm applied to the sequence set $\mathrm{S16}_{144}$. The plot shows the MC evolution of the sequence $\sbar$ (numbered 1--144), the total potential energy $E$ and the root-mean-square deviation (RMSD) calculated against the fold IA (light blue) and fold IB (dark red) structures shown in (B). The simulation is carried out at $\kb T = 0.43$ corresponding to the lowest studied temperature.  (B) Representative structures of fold IA and fold IB, selected to be the minimum-energy conformations found for the sequences A1 and N1, respectively. (C) Representative fold IIA and fold IIB structures selected to be the minimum-energy conformations of A2 and TN, respectively. }
\end{figure}

@@

\begin{figure}
\includegraphics[width=8.0cm]{Stderr}
\caption{Computational efficiency of the multisequence and simulated tempering algorithm.}
\end{figure}

We apply the multisequence algorithm to this set of 144 sequences across a range of temperatures. Figure~2 illustrates a typical simulation carried out at the lowest temperature, $\Tlow$, which is well below the folding temperature of both A1 and N1~\cite{Holzgrafe2014,Holzgrafe2015}. From the MC evolution of the total energy $E$, sequence index j, and similarity measures $Q_A$ and $Q_B$, it is evident that visits to various sequences drive transitions into a range of structural states. For example, visits to A1 ($\mathrm{j}=1$) and N1 ($\mathrm{j}=144$) tend to coincide with visits to high-$Q_A$ and high-$Q_B$ conformations, respectively, as is necessary to generate the correct conformational ensembles for these sequences.

Sequence updates $\sbar\rightarrow\sbar'$ are carried out by picking a new sequence $\sbar'$ randomly from the entire set of allowed sequences. Therefore, a sequence update typically changes more than one amino acid position. In other words, if the step size in sequence space is $\Delta H$, then typically $\Delta H>1$. Interestingly, despite the large $\Delta H$ and low $T$, the overall (average) acceptance rate $P_\mathrm{acc}=0.xx$ for $\sbar\rightarrow\sbar'$ updates which is not unreasonably low. Higher $P_\mathrm{acc}$ can be obtained by restricting proposed moves such that $\Delta H\le\Delta H_\mathrm{max}$, as shown in Figure~3. For example, an acceptance rate close to the oft-quoted rule-of-thumb value $0.25~\cite{Gilks1996} $ is obtained with $\Delta H_\mathrm{max}=4$, and with $\Delta H_\mathrm{max} = 1$ (only point mutations), $P_\mathrm{acc}=0.xx$. For simplicity, in this work, all simulations are carried without restrictions on how new sequences $\sbar'$ are drawn.

@@

To assess computational efficiency, we compare with results from simulated tempering (ST) simulations carried out on 4 of the 144 sequences, namely A1 and N1 and the two sequences R1 and R2 chosen randomly at $H=4$ and $H=6$ from A1, respectively (see Table~1). While ST provides the thermodynamics of a given sequence across the 8 temperatures $T_1, ..., T_8$ in a single run, a single MS simulation provides the thermodynamics of all 144 sequences for one $T_\mathrm{i}$. We adjust the simulation lengths for ST and MS runs such that an identical number of sampled conformations are obtained for each $\sbar$ and $T_\mathrm{i}$ combination, ensuring also that similar computational resources are used for the two algorithms.

Figure 4 shows the statistical error, $\sigE$, estimated for the average energy, $\left <E\right >$, for the 4 different sequences.

\begin{figure*}
\rotatebox{0}{\includegraphics[width=16.6cm]{ntwk16}}\\
\rotatebox{0}{\includegraphics[width=12.6cm]{ntwk35}}
\caption{Networks of the A1-N1 (top) and A2-TN (bottom) binary sequence space. }
\end{figure*}

@@

\vspace{12pt}
\subsection{A1-N1 and A2-TN binary sequence spaces}
\noindent
We now apply the MS algorithm to the sets {\SI} and {\SII} which represent the full binary sequence spaces for the A1-N1 and A2-TN systems, respectively. Both sets include $2^{10}=1,024$ sequences. Using MS simulations, we determined the thermodynamic behavior at the temperatures $\TlowI$ and $\TlowII$ for the A1-N1 and A2-TN systems, respectively (see Table~1). In particular, this allow us to find the stabilities of folds IA and IB ($\PaI$ and $\PbI$) for all sequences in {\SI} and the stabilities of folds IIA and IIB ($\PaII$ and $\PbII$) for all sequences in {\SII}.

Having calculated these fold stabilities, we are in a position to determine if there are pathways in sequence space that lead to IA-IB or IIA-IIB fold changes without passing through an unstable intermediate sequence. To this end, we construct graphs in which each stable sequence is represented by a node and any two nodes are connected if their sequences differ at only one amino acid position. To determine if a sequence is stable we used the criterion $P_\mathrm{tot}>\Pcut$, where  $\Ptot = \PaI + \PbI$ and $\PaII + \PbII$ for A1-N1 and A2-AN1, respectively. The precise network obtained depends, of course, on the cut-off value of $\Pcut$. In general, a higher $\Pcut$ means a selection of more stable pathways. Fig.~5 shows the networks obtained with $\Pcut=0.50$. The number of viable mutational pathways that connect the start and end points are 516,972 between A1 and N1 and  57,912 between A2 and AN1. Because there are $10! =3,628,800$ possible mutational pathways between start and end points in both cases, these pathways represent 14.2 \% and 1.6 \% of all possible pathways in A1-N1 and A2-AN1, respectively. Hence, A1 and N1 are rather highly connected at this stability threshold. For $\Pcut=0.60$, the numbers are 104,640 pathways (2.9\%) for A1-N1 and 22,512 (0.6\%) pathways for A2-AN1. We find that there are no possible A1-to-N1 or A2-to-AN1 pathways when $\Pcut\ge0.74$ and $\Pcut\ge 0.66$, respectively. \\

\subsection{Stability properties of fold-to-fold mutational pathways}
\noindent 
An apparently general characteristic of designed and natural protein folding switching is that the stability of sequences near the switch point have reduced stability.~\cite{Bryan2010} The results from our model exhibit a similar trend. Fig.~6A and B shows the total stability $\Ptot$ at different mutational distances between start and end points. Intermediate sequences are less stable than at distances $h=0$ and $h=10$, although there are large variations as indicated by the spread between upper and lower bounds. Nonetheless, there is a clear statistical trend that sequences become gradually less stable as successive mutations are applied to any of the 4 starting points until a minimum is reached. 
%These minima in stability occur closer to N1 and TN in the two systems, respectively, indicating that the IA/IIA folds are more mutationally robust than the IB/IIB folds.

\begin{figure}
\includegraphics[width=7.8cm]{Paths}
\caption{Averages stability as a function of Hamming distance $h$. }
\end{figure}

@@

For comparison, we used the well-established generalized-ensemble ST algorithm and focused on 144 model sequences with 16 amino acids ({\SR}). As a measure of computational efficiency, we used on the statistical error $\sigE$ in the estimation of the average total energy $E$ for 4 different sequences which are part of {\SR}.

@@

\end{document}

>>>>

\begin{abstract}
We evaluate a generalized-ensemble algorithm for coarse-grained simulations of biomolecules that allows the thermodynamics of multiple sequences to be calculated in a single ``multisequence" run. By carrying out a random walk in sequence space, conformational sampling is also enhanced. Escape from local energy minima is achieved by visiting sequences for which the minima is shallower or absent. We apply the algorithm alongside a coarse-grained model for protein folding with 7 atoms per amino acid to sets of sequences with up to 35 amino acids spanning a large sequence space. With the multisequence algorithm, we show that the folding thermodynamics of $>$ 1,000 sequences at relatively low temperatures can be readily calculated. We use the resulting data to explore the statistics of pathways in sequence space between two pairs of folds.

@@

\section{Introduction}
\noindent
Modern investigations into biophysical properties of biomolecular systems are routinely carried out not only on a single species but on multiple molecular variants.~\cite{Vukmirovic2000,Nickson2010} For example, protein engineering methods are used to provide mechanistic insight into the specificity of protein-protein and protein-DNA interactions, to compare the  conformational propensities of disease-related mutations of aggregation-prone proteins, and optimize biomolecular properties through directed evolution. Moreover, large-scale screening techniques such as phage display and microarray chips are used to provide biophysical properties across entire genomes. In short, through the study of a biomolecular system and related variants, molecular variants are studied in order to gain a more complete picture of the biophysical and biochemical properties of the system and provide evolutionary context.

@@

We first apply the MS algorithm to a set of 144 model sequences with 16 amino acids and compare with the simulated tempering (ST) algorithm. We find that the computational efficiency of the MS algorithm is never less than ST except at the lowest studied temperature for one sequence. The model folds spontaneously designed sequences into stable native states with diverse protein-like structures, and captures features such as mutational robustness. It therefore provides a biophysical basis for the sequence-structure and sequence-stability relationships of proteins at a conceptual level. This model has been used, e.g., to study the sequence-structure relationship of proteins in the "marshlands" between protein folds. Experimental studies have indicated that single point mutations are capable of changing one protein fold into another. The multisequence algorithm allows us to characterize the stability and structure of all 1,024 sequences in the binary space between the two folds. The implication of these results for the fold switching and evolution is discussed. \\

@@

While the 16-amino acid A1 and N1 sequences fold into $\alpha$-helical and $\beta$-hairpin folds, respectively, the two 35-amino acid sequences A2 and TN adopt the folds denoted IIA and IIB, as shown in Fig.~X.

The large number of sequences considered allows us to carry out a much more systematic analysis than previously~\cite{Holzgrafe2014,Holzgrafe2015} of the sequence-structure-stability relationship in the sequence region that connect these two folds pairs. We find that sequences become gradually less stable as they approach switch point. However, fold switching is observed to be typically abrupt at the switch point. Folding pathways usually pass through bistable sequences which populate both folds up to a significant level. We show that while bistable sequences become less when selecting more stable pathways for A1-N1 system, they generally remain the same for the A2-TN system.

@@

\subsection{Generalized-ensemble algorithms and simulated tempering}
\noindent
Conventional Monte Carlo simulations of the canonical distribution is problematic at low temperatures $T$ for many physical systems because simulations tend to become trapped in local energy minima and hamper representative sampling of configurational space. The basic idea of generalized-ensemble algorithms~\cite{Mitsutake2001} is to alleviate this trapping problem by sampling states using a non-Boltzmann weight factor and/or expand the state space with additional dynamical parameters, such that a more efficient random walk in potential energy can be achieved.

@@

In analogy with ST, the MS algorithm simulates the joint probability distribution
\begin{equation}
P(\sbar,\rbar) =\dfrac{1}{\tilde{Z}}e^{-\beta E(\sbar,\rbar) + h(\sbar)}\,, 
\label{MS}
\end{equation}
where  
\begin{equation}
\tilde{Z} = \sum_{\sbar}\sum_{\rbar} e^{-\beta E(\sbar,\rbar)+ h(\sbar)}\,
\end{equation}
and the first sum goes over a set of allowed sequences $\sbar$. The simulation parameters $h(\sbar)$, similar to the parameters $g_m$ in ST, control the marginal distribution $P(\sbar)=\tilde{Z}^{-1}\sum_{\rbar} e^{-\beta E(\sbar,\rbar)+ h(\sbar)} = \tilde{Z}^{-1}e^{-\beta F(\sbar)+ h(\sbar)}$ and a roughly flat $P(\sbar)$ can be achieved by choosing $h(\sbar) \approx \beta F(\sbar)$, where $F(\sbar)$ is the free energy of sequence $\sbar$ at temperature $T$.

@@

\section{Model and Methods}
\subsection{Coarse-grained 3-letter model for protein folding}
\noindent
All calculations were carried out using the coarse-grained model for protein folding developed in Ref.~\citenum{Bhattacherjee2012}. In this model, there are 3 different amino acid types: hydrophobic (h), polar (p) and turn-type (t). The backbone chain is represented atomistically by the N, H, $\mathrm{C}_\alpha$, $\mathrm{H}_{\alpha 1}$, C$'$ and O atoms. By contrast, the sidechain represention is simplified to a single enlarged $\mathrm{C}_\beta$ atom, which is geometrically identical for h and p types. The sidechain is absent for the t type which instead has an $\mathrm{H}_{\alpha 2}$ atom. The t type is therefore closely related to glycine. All bond lengths, bond angles, and peptide plane angles (180$^\circ$) are held fixed. Hence, an $N$-amino acid chain conformation $\rbar$ can, for any sequence $\sbar$, therefore be described by the set of 2$N$ backbone torsional angles $\{\phi_i$, $\psi_i\}_{i=1}^{N}$.

This geometrical description is paired with a simplified but finely tuned energy function with 4 terms: $E= E_\mathrm{ev} + E_\mathrm{loc} + E_\mathrm{hb} + E_\mathrm{hp}$. The first two, $E_\mathrm{ev}$ and $E_\mathrm{loc}$, represent excluded-volume effects of all atoms and local electrostatic effects, respectively. The hydrogen-bond energy, $E_\mathrm{hb}$, represents directionally dependent interactions between NH and CO groups which are necessary for secondary structure formation. Finally, the ``hydrophobicity" term, $E_\mathrm{hp}$, implements pairwise Lennard-Jones-like interactions between the $C_\beta$ atoms of h amino acids which are necessary for driving chain collapse during folding. Various model parameters, e.g., the strengths of hydrophobic attractions and hydrogen bonding, were determined based on the ability of the model to spontaneously fold a set of model sequences with 18-54 amino acids into  structurally diverse and thermodynamically stable native states with both $\beta$ and $\alpha$-structure. As it turned out, this strategy made the model robust enough to fold sequences designed to have mixed $\alpha$ and $\beta$ structures.

@@

\begin{table}
\caption{\label{tab1} List of model sequences of different lengths $N$ studied in this work.}
\begin{ruledtabular}
\begin{tabular}{lcr}
Name & $N$ & Sequence\footnote{In addition, we study two sets of 1,024 sequences derived from the A1-N1 and A2-TN pairs (see text) denoted $\mathrm{S16}_{1024}$  and $\mathrm{S35}_{1024}$, respectively, and a restricted set $\mathrm{S16}_{144}$ with 144 sequences taken from $\mathrm{S16}_{1024}$ ($\mathrm{S16}_{144}$ was also studied in Ref.~\protect\citenum{Holzgrafe2014}). } \\
\hline
A1 & 16 & pphpphhpphpphhpp \\
N1 & 16 & phphphpttphphphp \\
R1 & 16 & pphhphptthpphhpp\\
R2 & 16 & ppphphhtthhphppp\\
A2 & 35 & (A1)ttt(A1)\\
TN & 35 & (A1)ttt(N1)\\
\end{tabular}
\end{ruledtabular}
\end{table}

\subsection{Monte Carlo simulation parameters}
\noindent
Both ST and MS simulations are carried out with two types of conformational updates $\rbar\rightarrow\rbar'$: (1) a global pivot move (20\%) which randomly picks a $\phi_\mathrm{i}$ angle or $\psi_\mathrm{i}$ angle and assign a new value between $-\pi$ and $\pi$; and (2) a semi-local move (80\%) which turns the $\phi_\mathrm{i}$ and $\psi_\mathrm{i}$-angles of 4 consecutive amino acids in a coordinated manner.~\cite{Favrin2001} In MS simulations, sequence updates $\sbar\rightarrow\sbar'$ are carried out by randomly picking a new sequence $\sbar'\ne\sbar$ and applying the accept-reject criterion in Eq.~\ref{accrej}. A sequence move is attempted every $1,000$ MC steps while temperature updates $m\rightarrow m'$ are attempted every 100 steps. The simulations carried out in this work are summarized in Tables~II.

\begin{table}
\caption{\label{tab2} List of simulations carried out in this work. }
\begin{ruledtabular}
\begin{tabular}{lccccr}
Runs & Algorithm & $\kb T$  & MC steps/run &  Sequences\\
\hline
32 & ST & 0.43--0.65 & $8\times 10^6$ &A1\\ 
32 & ST & 0.43--0.65 & $8\times 10^6$ &N1\\ 
32 & ST & 0.43--0.65 & $8\times 10^6$ &R1\\ 
32 & ST & 0.43--0.65 & $8\times 10^6$ &R2\\ 
32$\times$8\footnote{32  runs per temperature for 8 temperatures in the range given.} & MS &0.43--0.65& $144\times 10^6$ & $\mathrm{S16}_{144}$\\
16 & MS & 0.43  & $5\times 10^9$ &  $\mathrm{S16}_{1024}$ \\
16 & MS & 0.46 & $4\times 10^9$ &  $\mathrm{S35}_{1024}$ \\
\end{tabular}
\end{ruledtabular}
\end{table}

@@

For folds A and B, $Q$ is determined by defining a set of backbone-backbone hydrogen bonds that are present in those folds (ignoring N and C terminal amino acids). For fold A, the fold-specific set of contacts (i,j) between the CO group of amino acid i and the NH group of amino acid j is $\{$(2,6), (3,7), (4,8), (5,9), (6,10), (7,11), (8,12), (9,13), (10,14), (11,15)$\}$, and for fold B the contact set is $\{$(3,14), (5,12), (7,10), (10,7), (12,5), (14,3)$\}$. A hydrogen bond is considered formed if (i) the NHO and COH angles are both $>90^\circ$ and (ii) the HO distance is $<2.7$~{\AA}.

@@

\begin{figure}
\includegraphics[width=10.0cm]{MCTrajFolds}
\caption{(A) Example simulation trajectory of the MS algorithm applied to the sequence set $\mathrm{S16}_{144}$. The plot shows the MC evolution of the sequence $\sbar$ (numbered 1--144), the total potential energy $E$ and the root-mean-square deviation (RMSD) calculated against fold IA (light blue) and fold IB (dark red) structures shown in (B). The simulation is carried out at $\kb T = 0.43$ corresponding to the lowest studied temperature.  (B) Representative structures of fold IA and fold IB, selected to be the minimum-energy conformations found for the sequences A1 and N1, respectively. (C) Representative fold IIA and fold IIB structures selected to be the minimum-energy conformations of A2 and TN, respectively. }
\end{figure}

@@

\begin{figure}
\includegraphics[width=8.0cm]{Stderr}
\caption{Computational efficiency of the multisequence and simulated tempering algorithms for 4 sequences (A) A1, (B) N1, (C) R1 and (D) R2.}
\end{figure}

We apply the multisequence algorithm to this set of 144 sequences across a range of temperatures (see Table~II). Figure~3 illustrates a typical simulation carried out at the lowest temperature, $\Tlow$, which is well below the folding temperature of both A1 and N1~\cite{Holzgrafe2014,Holzgrafe2015}. From the MC evolution of the total energy $E$, sequence index $\sbar$, and similarity measures $Q_A$ and $Q_B$, it is evident that visits to various sequences drive transitions into a range of structural states. For example, visits to A1 ($\sbar=1$) and N1 ($\sbar=144$) tend to coincide with visits to high-$Q_A$ and high-$Q_B$ conformations, respectively, as is necessary to generate the correct conformational ensembles for these sequences.

Sequence updates $\sbar\rightarrow\sbar'$ are carried out by picking a new sequence $\sbar'$ randomly from the entire set of allowed sequences. Therefore, a sequence update typically changes more than one amino acid position. In other words, if the step size in sequence space is $\Delta H$, then typically $\Delta H>1$. Interestingly, despite the large $\Delta H$ and low $T$, the overall (average) acceptance rate $P_\mathrm{acc}=0.xx$ for $\sbar\rightarrow\sbar'$ updates which is not unreasonably low. Higher $P_\mathrm{acc}$ can be obtained by restricting proposed moves such that $\Delta H\le\Delta H_\mathrm{max}$, as shown in Figure~3. For example, an acceptance rate close to the oft-quoted rule-of-thumb value $0.25~\cite{Gilks1996} $ is obtained with $\Delta H_\mathrm{max}=4$, and with $\Delta H_\mathrm{max} = 1$ (only point mutations), $P_\mathrm{acc}=0.xx$. For simplicity, in this work, all simulations are carried out without restrictions on how new sequences $\sbar'$ are drawn.

@@

In order to assess computational efficiency, we compare with results from simulated tempering (ST) simulations carried out on 4 of the 144 sequences, namely A1 and N1 and the two sequences R1 and R2 chosen randomly at $H=4$ and $H=6$ from A1, respectively (see Table~I). While ST provides the thermodynamics of a given sequence across the 8 temperatures $T_1, ..., T_8$ in a single run, a single MS simulation provides the thermodynamics of all 144 sequences for one $T_\mathrm{i}$. We adjust the simulation lengths for ST and MS runs such that an identical number of sampled conformations are obtained for each $\sbar$ and $T_\mathrm{i}$ combination, ensuring also that similar computational resources are used for the two algorithms.

Figure 4 shows the statistical error, $\sigE$, estimated for the average energy, $\left <E\right >$, for the 4 different sequences using ST and MS algorithms. Since an identical number of sampled conformations were obatined for each s and $T_i$ combination, we can therefore compare the statistical errors directly. Both MS and ST algorithms give the same small error at the highest T studied which implies consistent result between them. It should be noted that the highest T studied is well above the folding temperature for all the 4 sequences. Therefore, this consistent result is not particularly surprising since at high-T the free-energy landscape is relatively smooth. Hence, conformational space requires little or no difficulty to sample.

There is a sequence-dependent statistical trend observed in Figure 6. As temperature decreases, the statistical error from the MS algorithm becomes smaller than those from ST. Even at the lowest T studied, below the folding temperature of all 4 sequences, we notice that the MS algorithm is at least consistent and most times better than ST in its estimates except for N1. This departure from the dominant trend is most likely due to statistical flunctuations, in which case, does not have any considerable significance.

Insight from the statistical measurements suggests that the MS algorithm has a similar computational efficiency at $high_T$ and even better at lower T's. The MS algorithm is able to achieve  better efficency by sampling a wider region of both conformational and sequence space. Sampling the sequence space in addition to conformational space ensures a decorrelated ensemble of states is generated.

\begin{figure*}
\rotatebox{0}{\includegraphics[width=16.6cm]{ntwk16}}\\
\rotatebox{0}{\includegraphics[width=12.6cm]{ntwk35}}
\caption{Networks of the A1-N1 (top) and A2-TN (bottom) binary sequence space showing the possible stable pathways from one fold to another. }
\end{figure*}

@@

\vspace{12pt}
\subsection{A1-N1 and A2-TN binary sequence space}
\noindent
We now apply the MS algorithm to the sets {\SI} and {\SII} which represent the full binary sequence space for the A1-N1 and A2-TN systems, respectively. Each set includes $2^{10}=1,024$ sequences. Using MS simulations, we determined the thermodynamic behavior at the temperatures $\TlowI$ and $\TlowII$ for the A1-N1 and A2-TN systems, respectively (see Table~1). In particular, this allows us to find the stabilities relative to folds IA and IB ($\PaI$ and $\PbI$) for all sequences in {\SI} and the stabilities relative to folds IIA and IIB ($\PaII$ and $\PbII$) for all sequences in {\SII}.

Having calculated these fold stabilities, we are in a position to determine if there are pathways in sequence space that lead to IA-IB or IIA-IIB fold changes without passing through an unstable intermediate sequence. To this end, we construct graphs in which each stable sequence is represented by a node and any two nodes are connected if their sequences differ at only one amino acid position. To determine if a sequence is stable we used the criterion $P_\mathrm{tot}>\Pcut$, where  $\Ptot = \PaI + \PbI$ and $\PaII + \PbII$ for A1-N1 and A2-TN, respectively. The precise network obtained depends, of course, on the cut-off value of $\Pcut$. In general, a higher $\Pcut$ means a selection of more stable pathways. Fig.~5 shows the networks obtained with $\Pcut=0.50$. The number of viable mutational pathways that connect the start and end points are 516,972 between A1 and N1 and  57,912 between A2 and TN. Because there are $10! =3,628,800$ possible mutational pathways between start and end points in both cases, these pathways represent 14.2 \% and 1.6 \% of all possible pathways in A1-N1 and A2-TN, respectively. Hence, A1 and N1 are rather highly connected at this stability threshold. For $\Pcut=0.60$, the numbers are 104,640 pathways (2.9\%) for A1-N1 and 22,512 (0.6\%) pathways for A2-TN. We find that there are no possible A1-to-N1 or A2-to-TN pathways when $\Pcut\ge0.74$ and $\Pcut\ge 0.66$, respectively. \\

\subsection{Stability properties of fold-to-fold mutational pathways}
\noindent 
An apparently general characteristic of designed and natural protein folding switching is that sequences near the switch point have reduced stability.~\cite{Bryan2010} The results from our model exhibit a similar trend. Fig.~6A and B show the total stability $\Ptot$ at different mutational distances between start and end points. Intermediate sequences are less stable than sequences at distances $h=0$ and $h=10$, although there are large variations as indicated by the spread between upper and lower bounds. Nonetheless, there is a clear statistical trend that sequences become gradually less stable as successive mutations are applied to any of the 4 start and end points until a minimum is reached. 
%These minima in stability occur closer to N1 and TN in the two systems, respectively, indicating that the IA/IIA folds are more mutationally robust than the IB/IIB folds.

\begin{figure}
\includegraphics[width=7.8cm]{Paths}
\caption{(Left) A1-N1 System. (Right) A2-TN System. (Top) Average stability as a function of Hamming distance $h$. (Middle) Fraction of pathways as a function of Swicth length $L_s$. (Bottom) Average stability as a function of Bistability $B$.}
\end{figure}

@@

For comparison, we used the well-established generalized-ensemble ST algorithm and focused on 144 model sequences with 16 amino acids ({\SR}). As a measure of computational efficiency, we used the statistical error $\sigE$ in the estimation of the average total energy $E$ for 4 different sequences which are part of {\SR}.

@@

\end{document}
\grid
\grid
